---
title: "Activity_3_GEOG_331"
author: "Jacob Watts"
date: "9/23/2020"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Activity 3

# Author: Jacob Watts

## QA/QC at the Bewkes Center

# Question 1: Describe how the accuracy of the relative humidity varies with temperature and humidity.

1. Relative humidity is generally accurate within +/- 3 % RH. However, it does not exhibit uniform accuracy across the entire range of humidities and temperatures. It is most accurate at temperatures around 25 degrees Celcius and RH of 50%, where it exhibits +/- 2 % humidity. It is less accurate at the extremes of 0 or 100 % RH and 0 or 80 degrees C.

# Question 2: Describe the quality assurance protocols for the anemometer.
What unreliable measurements do they prevent? What are cases where data could be incorrectly marked as unreliable?

2. The anemometer takes a point measurement every 10 seconds and keeps a running average of the previous 10 measurements. The running average is used to determine if a measurement is erroneous or not. If a single measurement is more than 8 times the running average, the measurement is marked as rejected. This inbedded QA/QC prevents things such as power surges or animals messing with the anemometer readings. This could incorrectly mark a measurement as unreliable if a large front comes through or if the anemometer is in a wind tunnel.

# Question 3

```{r, include = T}
#read in the data file
#skip the first 3 rows since there is additional column info
#specify the the NA is designated differently
datW <- read.csv("bewkes_weather.csv",
                na.strings=c("#N/A"), skip=3, header=FALSE)
#preview data
print(datW[1,])

#get sensor info from file
# this data table will contain all relevant units
sensorInfo <-   read.csv("bewkes_weather.csv",
                na.strings=c("#N/A"), nrows=2)

print(sensorInfo)

#get column names from sensorInfo table
# and set weather station colnames  to be the same
colnames(datW) <-   colnames(sensorInfo)
#preview data
print(datW[1,])
```

The difference between skip and nrows is that skip tells R to skip a certain number of rows before reading the rest of the data while nrows tells R to read the first set of rows that you specify. So skip is used when multiple rows are used for the header as in this dataset. And nrows can be used to assign the names of the columns in the true dataset. These are quite different functions. The Header=F acts to tell R that the first line is not the header and to start reading the data right away from the first line. Since we skipped the header it was appropriate to set header to FALSE. The default is TRUE.

# Question 4

```{r, include = T}
#use install.packages to install lubridate
#install.packages(c("lubridate"))
library(lubridate)

#convert to standardized format
#date format is m/d/y
dates <- mdy_hm(datW$timestamp, tz= "America/New_York")

#calculate day of year
datW$doy <- yday(dates)
#calculate hour in the day
datW$hour <- hour(dates) + (minute(dates)/60)
#calculate decimal day of year
datW$DD <- datW$doy + (datW$hour/24)
#quick preview of new date calcualtions
datW[1,]

#see how many values have missing data for each sensor observation
#air temperature
length(which(is.na(datW$air.temperature)))

#wind speed
length(which(is.na(datW$wind.speed)))

#precipitation
length(which(is.na(datW$precipitation)))

#soil temperature
length(which(is.na(datW$soil.moisture)))

#soil moisture
length(which(is.na(datW$soil.temp)))

#make a plot with filled in points (using pch)
#line lines
plot(datW$DD, datW$soil.moisture, pch=19, type="b", xlab = "Day of Year",
     ylab="Soil moisture (cm3 water per cm3 soil)")

# there are missing values in the soil plot so we should ask the data gatherers what happened here.

#make a plot with filled in points (using pch)
#line lines
plot(datW$DD, datW$air.temperature, pch=19, type="b", xlab = "Day of Year",
     ylab="Air temperature (degrees C)")

#the first argument is a logical statement to be evaluated as true or false on a vector
#the second argument is the value that my air.tempQ1 column will be given if the statement
#is true. The last value is the value that will be given to air.tempQ1 if the statement is false.
#In this case it is just given the air temperature value
datW$air.tempQ1 <- ifelse(datW$air.temperature < 0, NA, datW$air.temperature)

#check the values at the extreme range of the data
#and throughout the percentiles
quantile(datW$air.tempQ1)

# let's visualize the really low data values

#look at days with really low air temperature
datW[datW$air.tempQ1 < 8,]  

 #look at days with really high air temperature
datW[datW$air.tempQ1 > 33,]  

```

The extremely low air temperature values are close to those values found on weatherunderground.com so it seems reasonable that the really low values were measured properly by the sensor. The extremely high air temperatures are also consistent with those reported by weatherunderground, which leads me to believe that the sensor was accurate in its measurements. Some dicrepancies are to be expected as the two sources are drawing data from two different locations.

# Question 5

```{r, include = T}

#plot precipitation and lightning strikes on the same plot
#normalize lighting strikes to match precipitation
lightscale <- (max(datW$precipitation)/max(datW$lightning.acvitivy)) * datW$lightning.acvitivy
#make the plot with precipitation and lightning activity marked
#make it empty to start and add in features
plot(datW$DD , datW$precipitation, xlab = "Day of Year", ylab = "Precipitation & lightning",
        type="n") + points(datW$DD[datW$precipitation > 0], datW$precipitation[datW$precipitation > 0],
        col= rgb(95/255,158/255,160/255,.5), pch=15) + points(datW$DD[lightscale > 0], lightscale[lightscale > 0],
        col= "tomato3", pch=19)

# why can lightscale object be used to subset datW even if it not part of datW?

assert <- function(statement,err.message){
  #if evaluates if a statement is true or false for a single item
  if(statement == FALSE){
    print(err.message)
  }
  
}

#check how the statement works
#evaluate a false statement
assert(length(lightscale) == length(datW$precipitation), "error: unequal values")

# assert yields no error message. The two vectors are the same length.
```

Lightscale can be used to subset values in datW for two reasons. For one, lightscale was created by doing simple math of multiple columns within datW. Therefore its values are directly related to precipitation and lightning. Second, the assert function demonstrates the lightscale is the same length as datW$precipitation. This is another requirement for the subset function. 

# Question 6

```{r, include = T}
#filter out storms in wind and air temperature measurements
# filter all values with lightning that coincides with rainfall greater than 2mm or only rainfall over 5 mm.    
#create a new air temp column
datW$air.tempQ2 <- ifelse(datW$precipitation  >= 2 & datW$lightning.acvitivy >0, NA,
                    ifelse(datW$precipitation > 5, NA, datW$air.tempQ1))
names(datW)
# same for wind speed
datW$wind.speedQ1 <- ifelse(datW$precipitation  >= 2 & datW$lightning.acvitivy >0, NA,
                    ifelse(datW$precipitation > 5, NA, datW$wind.speed))


# proving that this worked

assert <- function(statement,err.message){
  #if evaluates if a statement is true or false for a single item
  if(statement == FALSE){
    print(err.message)
  }
  
}
# new vectors which remove NA
wind <- (datW$wind.speedQ1)[!is.na(datW$wind.speedQ1)]
airtemp <- (datW$air.tempQ2)[!is.na(datW$air.tempQ2)]

#check how the statement works
#evaluate a false statement
assert(length(wind) == length(airtemp), "error: unequal values")

# no error message.
plot(datW$DD, datW$wind.speedQ1, pch=19, type="b", xlab = "Day of Year",
     ylab="Wind Speed (m/sec)")

```

There were 13 measurements that were deemed to be unreliable because of their association with lightning and precipitation. This is relatively little considering that there are 2118 points. This sensor seems to be very reliable. I knew that my filter function worked by removing the NAs and counting how many NAs were removed using the assert function. My assert function yielded no error message.

# Question 6

```{r, include = T}
# Sudden changes in the sensor position in the ground are marked by dramatic changes in soil temperature and moisture.

names(datW)

plot(datW$DD, datW$soil.moisture, pch=19, type="b", xlab = "Day of Year",
     ylab="Soil moisture (cm3 water per cm3 soil)") + points(datW$DD[datW$precipitation > 0], datW$precipitation[datW$precipitation > 0],
        col= rgb(95/255,158/255,160/255,.5), pch=15)
  

# cyclical soil moisture pattern, follows the rain
plot(datW$DD, datW$soil.temp, pch=19, type="b", xlab = "Day of Year",
     ylab="Soil Temp (degrees C))") + points(datW$DD[datW$air.tempQ2 > 0], datW$air.tempQ2[datW$air.tempQ2 > 0],
        col= rgb(95/255,158/255,160/255,.5), pch=15)
# collinearity
cor.test(datW$soil.temp, datW$air.temperature)

# temperature correlated with the soil temperature with a correlation value of 0.947
  
```

Based on a visual analysis of the graph of soil moisture and rainfall, it looks like the soil moisture spikes when it rains and slowly depletes from there. There do not seem to be any outliers in the data. This indicates to me that the sensor was not moved before it was disconnected. The graph of soil temperature and air temperature show that both are highly correlated, with no obvious outliers. Also the correlation test between the two variables shows that they are highly correlated with a value of 0.947. I am confident that the available data from the soil probe are acceptable up until the point that it was disconnected.

# Question 7

```{r, include = T}
library(expss)
# making a table with labels. average air temp, wind speed, soil temp, soil moist
datW = apply_labels(datW,
                      wind.speedQ1 = "Wind Speed (m/s)",
                    air.tempQ1 = "Air Temperature (degrees C)",
                    soil.temp = "Soil Temperature (degrees C)",
                    soil.moisture = "Soil Moisture (m^3 h20/m^3 soil")
datW %>% 
    tab_cells(wind.speedQ1, air.tempQ1, soil.temp, soil.moisture) %>%
    tab_cols(total(label = "Average Climatic Conditions| |")) %>% 
    tab_stat_fun(Mean = w_mean, "Std. dev." = w_sd, "Valid N" = w_n, method = list) %>%
    tab_pivot() %>% 
    set_caption("Table with summary statistics of weather in Hamilton NY.")
# total precipitation
total_precip <- sum(datW$precipitation)
total_precip
# Study period
max(datW$DD) - min(datW$DD)
# dates
datW$timestamp [1]
datW$timestamp [2118]
```

The above table indicates the mean and standard deviation of the values requested by the researchers. The number of observations is indicated by the Valid N column. The total precipitation over the study period was 177.828 mm. The study was conducted from 6/12/2018 11:30 to 7/26/2018 14:00, a 44 day period.

# Question 9

```{r, include = T}
par(mar = c(4, 4, .1, .1))
plot(datW$DD, datW$soil.temp, pch=19, type="b", xlab = "Day of Year",
     ylab="Soil Temp (degrees C))")
plot(datW$DD, datW$air.temperature, pch=19, type="b", xlab = "Day of Year",
     ylab="Air temperature (degrees C)")
plot(datW$DD, datW$precipitation, pch=19, type="b", xlab = "Day of Year",
     ylab="Precipitation (mm)")
plot(datW$DD, datW$soil.moisture, pch=19, type="b", xlab = "Day of Year",
     ylab="Soil Moisture (m^3 h20/m^3 soil")
```

The air temperature follows a diurnal pattern, with colder nights and warmer days. There also appears to be larger cyclical pattern that follows the ebs and flows with High and Low pressure systems and weather events. The soil temperature closely follows the air temperature, though it is generally a degree or two lower than the air temperature likely due to evaporative cooling and shading from plant foliage. Precipitation is highly sporadic, with 8 distinct prepitation events across the 44 day study period. The soil moisture rapidly increases after precipitation events and falls until the next precipitation event.

# Question 10

https://github.com/jlwatts98/GEOG331/blob/master/actvity_3.Rmd


